Laurens de Graaf

### 2015-05-06 writeup/paper.tex
Write up is done

\begin{abstract}
\begin{center}
\emph{``A Great and Mischievous Pirate''
{\begin{flushright}$\sim$ Henry Morgan\end{flushright}}}
\end{center}
\end{abstract}

@@

def loop(config):
  while not terminateHuh(config):
    jitdriver.jit_merge_point(code = 
                                config.code)
    config = config.code.step(config)
  return config
\end{code}
\caption{\label{fig:loop}The main loop and termination test.}
\end{figure}

@@

\paragraph{\textit{ReturnInt i}.}
This operation indicates that an expression has evaluated to an integer $i$. The
return stack will contain a \lstinline{case} expression to match against this
result. We present this case in full in Figure~\ref{fig:ReturnInt}; the 
\lstinline{int_case_lookup} operator takes the relevant cases and identifies the
appropriate one. We also indicate a \emph{default case}, which uses a binder
with a default variable to bind the result if it does not match an integer
literal.

@@

\begin{code}
from rpython.rlib.jit import JitDriver, purefunction

@@

jitdriver = JitDriver(greens=['code']
                     , reds='auto'
                     , get_printable_location=get_location)

@@

  while not terminateHuh(config):
    jitdriver.jit_merge_point( code       = config.code)
    config = config.code.step(config)
  return config
\end{code}

@@

Finally, we discuss our timing results, and must report a negative one:
the final implementation is \emph{slower} when we perform tracing on
a sample program computing the 30th Fibonacci number. This is almost certainly
the result of our na\"{i}ve tracing mechanism, wherien we consider every node as
a possible merge point for the JIT, introducing immense tracing overhead to the
computation. Thus it is unclear if Tracing is suitable for the STG machine
because we have not performed effort to fine-tune our implementation.

Beyond this result, we still have much to do:
\begin{itemize}
\item Write a full parser that will construct AST nodes from reasonable input;
\item Write a more extensive test suite to examine and use constructors,
      \lstinline{let}, \lstinline{letrec}, laziness, etc.; this will be much
      easier once a parser is written;
\item Support a full range of built-in functions, including I/O and additional
      numeric operators;
\item Replace the dictionary-based representation of environments with a
      more efficient data structure
\item Fine-tune the JIT heuristics to gain speed.
\end{itemize}

@@

%% \acks
%% 
%% Acknowledgments, if needed.

>>>>

\begin{abstract}
\begin{center}
\emph{``A Great and Mischievous Pirate''
{\begin{flushright}$\sim$ Henry Morgan\end{flushright}}}
\emph{``I didn't proof-read this paper before I turned it in. I'm certain it's rough,
       and needs at least one more rewrite.''
{\begin{flushright}$\sim$ Cameron Swords\end{flushright}}}
\end{center}
\end{abstract}

@@

def loop(config):
  while not terminateHuh(config):
    config = config.code.step(config)
  return config
\end{code}
\caption{\label{fig:loop}The main loop and termination test.}
\end{figure}

@@

\paragraph{\textit{ReturnInt i}.}
This operation indicates that an expression has evaluated to an integer $i$. The
return stack will contain a \lstinline{case} expression to match against this
result. We present this case in full in Figure~\ref{fig:returnint}; the 
\lstinline{int_case_lookup} operator takes the relevant cases and identifies the
appropriate one. We also indicate a \emph{default case}, which uses a binder
with a default variable to bind the result if it does not match an integer
literal.

@@

We implement a na\"{i}ve JIT driver with RPthon in our main loop, modifying it
to support a merge point and identifying the \lstinline{code} piece of the 
configuration as the \emph{green} portion. This modified main loop is presented
in Figure~\ref{fig:jitloop}.
\begin{figure}
\begin{code}
from rpython.rlib.jit 
  import JitDriver, purefunction

@@

jitdriver = 
  JitDriver( greens=['code']
           , reds='auto'
           , get_printable_location =
               get_location)

@@

  while not terminateHuh(config):
    jitdriver.jit_merge_point( code = 
                                 config.code)
    config = config.code.step(config)
  return config
\end{code}
\caption{\label{fig:jitloop}The JIT-sensitive main loop.}
\end{figure}

\subsection{Timing Results}

Our timing results are surprisingly poor: 
the final implementation is \emph{slower} when we perform tracing on
a sample program computing the 30th Fibonacci number, as opposed to running
it through PyPy (at 18 seconds versus 14 seconds). Perhaps less shocking,
GHC can compile and run an equivalent program in approximately 0.022 seconds,
suggesting we are several \emph{orders of magnitude} slower than the current
GHC implementation.

This is almost certainly the result of our na\"{i}ve tracing mechanism, wherein
we consider every node as a possible merge point for the JIT, introducing
immense tracing overhead to the computation. The first optimization we should
explore is only considering \textit{Enter} operations for merge points. The
second, less clear optimization, is that of termination conditions: using a
termination condition as we have in our implementation can be hard on the
tracing mechanism, but it is unclear without further investigation how our STG
implementation might be modified to ``politely'' terminate, given the continuous
interaction of the return and update stacks.

@@

In closing, we found this project enlightening and insightful: the idea of a 
tracing JIT is appealing, but there are a number of difficulties and concerns
that arise in the area almost immediately that we have tried to discuss in this
paper. Tracing seems, in some ways, a brittle box---at least in RPython, there
are a number of hoops to leap through to get code compiled. To this end,
it would would be interesting to see a similar system built in a strongly-typed
language in order to avoid the ``surprise'' compilation errors that one
experiences when building facilities in RPython. Compared to the standard
approach of building a compiler, however, this alternative has a lot of promise
in terms of rapid prototyping, language evolution, and general lack of
compiler-specific expert knowledge. While the STG machine may have been a poor
choice for implementation (due to the authors' lack of familiarity with its
intricacies), using a system like RPython in the future for quick prototyping
is certainly reasonable and worthwhile.

Beyond this, we still have much to do:
\begin{itemize}
\item Write a full parser that will construct AST nodes from reasonable input;
\item Write a more extensive test suite to examine and use constructors,
      \lstinline{let}, \lstinline{letrec}, laziness, etc.; this will be much
      easier once a parser is written;
\item Support a full range of built-in functions, including I/O and additional
      numeric operators;
\item Replace the dictionary-based representation of environments with a
      more efficient data structure
\item Consider breaking up the \textit{Eval} implementation into individual
      AST nodes.
\item Fine-tune the JIT heuristics to gain speed.
\end{itemize}

@@

\acks

We'd like to thank Spenser Bauman for his immense interest and help
in this work, and Sam Tobin-Hochstadt for facilitating the creation
of this project as part of his Metaprogramming course.
